---
title: "Text Mining in R"
subtitle: "Searching and visually comparing different fields"
author: "Teresia Mrema-Buza"
date: "<strong>Last Updated</strong>: `r Sys.Date()`"
site: bookdown::bookdown_site
knit: bookdown::render_book
output:
  bookdown::bs4_book:
    includes:
      in_header:
    template: bs4_template.html
        
  bookdown::gitbook:
    includes:
      in_header: header.html
    config: 
      toc:
       collapse: 
      search: yes
      fontsettings:
        size: 2
    split_by: section
    split_bib: no
    df_print: paged
    number_sections: true
    css:
    - css/style.css
    - style.css
    
  html_document:
    code_folding: hide
    
  bookdown::html_document2:
    code_folding: hide
    
  bookdown::pdf_book:
    config:
      toc = true
      toc_depth = 2
      number_sections = true
      fig_caption = true
      keep_tex = no
      pandoc_args = null
      toc_unnumbered = no
      toc_appendix = no
      toc_bib = no 
      quote_footer = null
      highlight_bw = true
      latex_engine = xelatex
      df_print = kable
      base_format = rmarkdown::pdf_document
  includes:
    in_header: 
    - latex/header.tex
    - latex/preamble.tex
documentclass: book
classoption: openany #remove empty pages in pdf doc
bibliography:
- library/book.bib
- library/packages.bib
- library/microbiome.bib
- library/software.bib 
citation_package:
- natbib
- biblatex
- amsplain
colorlinks: true
css:
- css/style.css
always_allow_html: true
fig_caption: true
fontsize: 12pt
geometry: margin=1in
indent: false
keep_tex: true
link-citations: true
mainfont: Times New Roman
biblio-style: apalike
spacing: double
header-includes: 
- \usepackage{setspace}
- \newpage
- \newenvironment{tmbinfo}[0]{}{}
- \renewenvironment{tmbinfo}[0]{}{}
- \newenvironment{tmbalert}[0]{}{}
- \renewenvironment{tmbalert}[0]{}{}
- \newenvironment{tmbshare}[0]{}{}
- \renewenvironment{tmbshare}[0]{}{}
description: 
--

```{r setup, include=FALSE}
source("R/common.R")
library(flexdashboard)
library(tidyverse, suppressPackageStartupMessages())
```


# Pubmed Literature Tracking in R
```{r pubmed, fig.cap="PubMed Search Workflow"}
library(DiagrammeR)
library(DiagrammeRsvg)

mermaidgraph <- mermaid("graph TD
A(OMICS PUBLICATION STATISTICS) --> A0{NCBI PubMed Database}
A0 --> A1[Microbiome OR Metabolomics]
A0 --> A2[Genomics]
A0 --> A3[Proteomics]
A0 --> A4[Metabolomics]
A0 --> A5[Phenomics]
A0 --> A6[Transcriptomics]


subgraph ARTICLE SEARCH USING OMICS KEYWORDS
A1 --> B{Indexed Papers in PubMed}
A2 --> B
A3 --> B
A4 --> B
A5 --> B
A6 --> B

B --> |Download results| C[Results by Year in CSV format]
end

subgraph SEARCH RESULTS
C --> |Filename| D[PubMed_Timeline_Results_by_Year.csv]
D --> E[Save file in data directory]
E -.-> F[Further Analysis]
end

", height = 600, width = 600)

mermaidgraph

```



